Compositional Certified Resource Bounds

### 2015-04-14 paper/main.tex
add large table
mention the large table in section 8
trying to save some space, not enough

\usepackage[utf8]{inputenc} %for utf8 input
\usepackage{amssymb} %for shift symbol
\usepackage{amsmath}
\usepackage{listings} %for code
\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
\usepackage{mathpartir} %for typing rules
\usepackage{microtype} %better micro typing
\usepackage{stmaryrd} %for llbracket
\usepackage{mathabx} % for boxes
\usepackage{graphicx} %to include png images
\usepackage{xcolor} %for colors
\usepackage{url}
\usepackage{enumitem}
\usepackage{array} %for stupid tables
\usepackage{flushend} %equalize columns length on last page
%\usepackage[firstpage]{draftwatermark}
%\SetWatermarkText{\hspace*{8in}\raisebox{5.5in}{\includegraphics[scale=0.1]{fig/aec-badge-pldi.pdf}}}
%\SetWatermarkAngle{0}

@@

This idea is illustrated by the program in \pref{fig:xmplincaux}.
The parts of the code in blue are annotations that were added
to the original program text.  The top-level loop increments
a binary counter $k$ times.  A naive analysis of
the algorithm yields the quadratic bound $k \cdot N$.
However, the algorithm is in fact linear and its cost is bounded
by $2k + \#_1(a)$ where $\#_1(a)$ denotes
the number of one entries in the array $a$.  Since this
number depends on the heap contents, no tool available
for C is able to derive the linear bound.  However, it can be inferred
by our automated tool if a logical variable $na$ is introduced.
This logical variable is a reification of the
number $\#_1(a)$ in the program.  For example, on line 6 of the example
we are setting \lstinline{a[x]} to 0 and because of the condition we
know that this array entry was 1.  To reflect this change on
$\#_1(a)$, the logical variable $na$ is decremented.
Similarly, on line 8, an array entry which was 0 becomes 1, so
$na$ is incremented.  To complete the step 2 of the systematic
procedure described above, we must show that the extra assertion
\lstinline{na > 0} on line 5 is always true.  It
is done by proving inductively that $na = \#_1(a)$
and remarking that since \lstinline{a[x] == 1} is true, we must have
$\#_1(a) > 0$, thus the assertion \lstinline{na > 0} never fails.

@@

Another simple example is given in \pref{fig:xmplbsaux} where
a logarithmic bound on the stack consumption of a binary search
program is proved using logical variable annotations.  Once again,
annotations are in blue in the program text.  In this example,
to ease the proof of equivalence between the annotated program
and the original one, we use the inequality $lg >
\log_2(h-l)$ as an invariant.  This technique allows a simpler
proof because, when working with integer arithmetic, it is not always
the case that $\log_2(x-x/2) = \log_2(x)-1$.

@@

The translation of an assertion $(\Gamma; Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma; Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  The logical context $\Gamma$
is implicitly lifted to a quantitative assertion by mapping a state
$\state$ to 0 if $\Gamma(\state)$ holds and to $\infty$ otherwise.
\iffull{We also need to
translate the assumptions in function contexts of the automatic
analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
\forall z \, \vec v \, v . $
%
$$
(
  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
  \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
  )
$$}{}
%
These definitions let us translate the judgement
$J := B, R \vdash \htriple P S {P'}$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
\iffull{$$
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$}{%
$$
  \tr(J) := \tr(B); \tr(R) \vdash_L \htriple{\tr(P)} S {\tr(P')}.
$$}%
\noindent{}The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the automatic analysis]
  If $J$ is a judgement derived by the automatic analysis,
  then $\tr(J)$ is a quantitative Hoare triple derivable
  in the quantitative logic.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and maps each rule
of the automatic analysis directly to its counterpart in the
quantitative logic.  The most tricky parts are the translations of the
rules for increments and decrements and the rule {\sc Q:Weak} for
weakening because they make essential use of the algebraic properties
of our linear potential functions.
\iffull{ For instance we prove the following lemma to show the soundness
of the translation of {\sc Q:Weak}.

@@

\pref{tab:compar} summarizes the results of our experiments.  It shows
for each tool the number of derived bounds (\#bounds), the number of
asymptotically tight bounds (\#lin. bounds), the number of bounds with
the best constant factors in comparison with the other tools (\#best
bounds), and the number of examples that we were able to test with the
tool (\#tested).  Since we were not able to run the experiments for KoAT
and SPEED, we simply used the bounds that have been reported by the
authors of the respective tools.  The results show that our automatic
amortized analysis outperforms the existing tools on our example
programs.  However, this experimental evaluation has to be taken with
a grain of salt.  Existing tools complement $\toolname$
since they can derive polynomial bounds and support
more features of C.  We were particularly impressed by LOOPUS which is
very robust, works on large C files, and derives very precise bounds.
\iffull{We did not include the running times of the tools in the table since
all tested tools work very efficiently and need less then a second
on every tested example.}{}

@@

\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
\newcommand{\file}[2][c]{%
  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
\renewcommand{\max}[0]{{\rm mx}}
\renewcommand{\min}[0]{{\rm mn}}

@@

File &
\multicolumn{2}{|c|}{KoAT} &
\multicolumn{2}{|c|}{Rank} &
\multicolumn{2}{|c|}{LOOPUS} &
SPEED &
\toolname
\\
\hline

%--------------------------------------------
\hline \texttt{gcd.c} &

@@

%--------------------------------------------
\end{tabular}
\caption{
  Experimental evaluation comparing the bounds generated KoAT, Rank,
  LOOPUS, SPEED, and our automatic amortized analysis on several
  challenging linear examples. Results for KoAT and SPEED were extracted
  from previous publications because KoAT cannot take C programs as
  input in its current version and SPEED is not available.
  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
  marked with --- indicate that the tool failed to produce a result.
}
\label{tab:eval}
\end{table*}

>>>>

\usepackage[utf8]{inputenc} %for utf8 input
\usepackage{amssymb} %for shift symbol
\usepackage{amsmath}
\usepackage{listings} %for code
\lstset{moredelim=[is][\color{blue}]{@}{@},keepspaces=true}
\usepackage{mathpartir} %for typing rules
\usepackage{microtype} %better micro typing
\usepackage{stmaryrd} %for llbracket
\usepackage{mathabx} % for boxes
\usepackage{graphicx} %to include png images
\usepackage{xcolor} %for colors
\usepackage{url}
\usepackage{enumitem}
\usepackage{array} %for stupid tables
\usepackage{flushend} %equalize columns length on last page
\usepackage{hhline}
%\usepackage[firstpage]{draftwatermark}
%\SetWatermarkText{\hspace*{8in}\raisebox{5.5in}{\includegraphics[scale=0.1]{fig/aec-badge-pldi.pdf}}}
%\SetWatermarkAngle{0}

@@

This idea is illustrated by the program in \pref{fig:xmplincaux}.
The parts of the code in blue are annotations that were added
to the original program text.  The top-level loop increments
a binary counter $k$ times.  A naive analysis of
the algorithm yields the quadratic bound $k \cdot N$.
However, the algorithm is in fact linear and its cost is bounded
by $2k + \#_1(a)$ where $\#_1(a)$ denotes
the number of one entries in the array $a$.  Since this
number depends on the heap contents, no tool available
for C is able to derive the linear bound.  However, it can be inferred
by our automated tool if a logical variable $na$ is introduced.
This logical variable is a reification of the
number $\#_1(a)$ in the program.  For example, on line 6 of the example
we are setting \lstinline{a[x]} to 0 and because of the condition we
know that this array entry was 1.  To reflect this change on
$\#_1(a)$, the logical variable $na$ is decremented.
Similarly, on line 8, an array entry which was 0 becomes 1, so
$na$ is incremented.  To complete the step 2 of the systematic
procedure described above, we must show that the extra assertion
\lstinline{na > 0} on line 5 cannot fail.  We do it
by proving inductively that $na = \#_1(a)$
and remarking that since \lstinline{a[x] == 1} is true, we must have
$\#_1(a) {>} 0$, thus the assertion \lstinline{na > 0} never fails.

@@

Another simple example is given in \pref{fig:xmplbsaux} where
a logarithmic bound on the stack consumption of a binary search
program is proved using logical variable annotations.  Once again,
annotations are in blue in the program text.  In this example,
to ease the proof of equivalence between the annotated program
and the original one, we use the inequality $lg >
\log_2(h-l)$ as invariant.  This allows a simpler
proof because, when working with integer arithmetic, it is not always
the case that $\log_2(x-x/2) = \log_2(x)-1$.

@@

The translation of an assertion $(\Gamma; Q)$ in the automatic
analysis is defined by
$$
  \tr(\Gamma; Q) := \lambda\state .\, \Gamma(\state) + \Phi_Q(\state),
$$
where we write $\Phi_Q$ for the unique linear potential function
defined by the quantitative annotation $Q$.  The logical context $\Gamma$
is implicitly lifted to a quantitative assertion by mapping a state
$\state$ to 0 if $\Gamma(\state)$ holds and to $\infty$ otherwise.
\iffull{We also need to
translate the assumptions in function contexts of the automatic
analysis.  We define $\tr(\Gamma_f, Q_f, \Gamma'_f, Q'_f) :=
\forall z \, \vec v \, v . $
%
$$
(
  \lambda \vec v \, \state .\, \tr(\Gamma_f, Q_f)(\state[\Vargs {\mapsto} \vec v]),
  \lambda v \, \state .\, \tr(\Gamma'_f, Q'_f)(\state[\Vret {\mapsto} v])
  )
$$}{}
%
These definitions let us translate the judgement
$J := B, R \vdash \htriple P S {P'}$\iffull{ in the context $\Delta$ by $\tr(J) :=$}{ by}
\iffull{$$
  (\lambda f .\, \tr(\Delta(f))); \tr(B); \tr(R) \vdash
    \htriple{\tr(P)}{S}{\tr(P')}.
$$}{%
$$
  \tr(J) := \tr(B); \tr(R) \vdash_L \htriple{\tr(P)} S {\tr(P')}.
$$}%
\noindent{}The soundness of the automatic analysis can now be stated
formally with the following theorem.
%
\begin{theorem}[Soundness of the automatic analysis]
  If $J$ is a judgement derived by the automatic analysis,
  then $\tr(J)$ is a quantitative Hoare triple derivable
  in the quantitative logic.
\end{theorem}
%
\noindent
The proof of this theorem is constructive and maps each rule
of the automatic analysis directly to its counterpart in the
quantitative logic.  The trickiest parts are the translations of the
rules for increments and decrements and the rule {\sc Q:Weak} for
weakening because they make essential use of the algebraic properties
of the potential functions.
\iffull{ For instance we prove the following lemma to show the soundness
of the translation of {\sc Q:Weak}.

@@

\pref{tab:compar} summarizes the results of our experiments presented
in \pref{tab:eval}.  It shows
for each tool the number of derived bounds (\#bounds), the number of
asymptotically tight bounds (\#lin. bounds), the number of bounds with
the best constant factors in comparison with the other tools (\#best
bounds), and the number of examples that we were able to test with the
tool (\#tested).  Since we were not able to run the experiments for KoAT
and SPEED, we simply used the bounds that have been reported by the
authors of the respective tools.  The results show that our automatic
amortized analysis outperforms the existing tools on our example
programs.  However, this experimental evaluation has to be taken with
a grain of salt.  Existing tools complement $\toolname$
since they can derive polynomial bounds and support
more features of C.  We were particularly impressed by LOOPUS which is
very robust, works on large C files, and derives very precise bounds.
\iffull{We did not include the running times of the tools in the table since
all tested tools work very efficiently and need less then a second
on every tested example.}{}

@@

\newcommand{\dir}[2]{\texttt{#1} \\ \texttt{#2}}
\newcommand{\file}[2][c]{%
  \begin{tabular}[#1]{@{}r@{}}#2\end{tabular}}
\renewcommand{\max}[0]{{\rm mx}}

@@

File &
\multicolumn{2}{c|}{KoAT} &
\multicolumn{2}{c|}{Rank} &
\multicolumn{2}{c|}{LOOPUS} &
SPEED &
\toolname
\\

%--------------------------------------------
\hhline{=|==|==|==|=|=} \texttt{gcd.c} &

@@

%--------------------------------------------
\end{tabular}
\caption{
  Experimental evaluation comparing the bounds generated KoAT, Rank,
  LOOPUS, SPEED, and our automatic amortized analysis on several
  challenging linear examples. Results for KoAT and SPEED were extracted
  from previous publications because KoAT cannot take C programs as
  input in its current version and SPEED is not available.
  Entries marked with ? indicate that we cannot test the respective example with the tool. Entries
  marked with --- indicate that the tool failed to produce a result. We write
  $\max(a,b)$ for the maximum of $a$ and $b$.
}
\label{tab:eval}
\end{table*}
